COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-23 TACO-paper/coffee.tex
Fix layout and style and names of algos/listings

\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfig}
\usepackage{tabulary}
\usepackage[cmex10]{amsmath}
\usepackage{url}

@@

\begin{algorithm}[t]
\footnotesize
\KwIn{element matrix (2D array, initialized to 0),\\~~~~~~~~~element coordinates (array),\\~~~~~~~~~coefficient fields (array, e.g. velocity)}
\KwOut{element matrix (2D array)}
~~- Compute Jacobian from coordinates \\
~~- Define basis functions \\
~~- Compute element matrix in an affine loop nest
\caption{General structure of a local assembly kernel generated by Firedrake.}
\label{code:general-structure}
\end{algorithm}

@@

\begin{algorithm}[t]
\footnotesize
\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$3; k++) \\
~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
~~~~~~~~~~~+((K1*X$\_$D10[i][k]+K3*X$\_$D01[i][k])*(K1*X$\_$D10[i][j]+K3*X$\_$D01[i][j]))+\\
~~~~~~~~~~~+((K0*X$\_$D10[i][k]+K2*X$\_$D01[i][k])*(K0*X$\_$D10[i][j]+K2*X$\_$D01[i][j])))*\\
~~~~~~~~~~~*det*W[i]);\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for a Helmholtz problem on a 2D triangular mesh using Lagrange $p=1$ elements.}
\label{code:helmholtz}
\end{algorithm}

\begin{algorithm}[t]
\footnotesize
\KwSty{void} burgers(double A[12][12], double **coords, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//11 other basis functions definitions.\\
~~...\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[i][r]);\\
~~~~~~//10 analogous statements (F1, F2, ...)\\
~~~~~~...\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K5*F9)+(K8*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K0*X1$\_$D100[i][k])+(K3*X1$\_$D010[i][k])+(K6*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K2*X2$\_$D100[i][k])+...+(K8*X2$\_$D001[i][k]))*((K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$roughly a hundred sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W[i]);\\
~~$\rbrace$ \\
$\rbrace$
\caption{Local assembly code generated by Firedrake for a Burgers problem on a 3D tetrahedral mesh using Lagrange $p=1$ elements.}
\label{code:burgers}
\end{algorithm}

@@

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions. An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse. In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure.

@@

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens of iterations, render the optimization process challenging. It is not always the same set of optimizations that bring performance closest to the machine peak. For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces. Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.

@@

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From the inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions which only depend on two iteration variables. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $intel$ and $gnu$ compilers exhibit. First, they only identify sub-expressions that are invariant with respect to the innermost loop. This is an issue for sub-expressions depending on \texttt{i}-\texttt{k}, which are not automatically lifted in the loop order \texttt{ijk}. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization.

@@

Figure~\ref{code:helmholtz-licm} shows the Helmholtz assembly code after the application of loop-invariant code motion, padding, and data alignment.

\begin{algorithm}[t]
\footnotesize
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4] ALIGN;\\
~~~~double LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+LI$\_$1[k]*LI$\_$1[j])*det*W[i]);\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code for the Helmholtz problem in Figure~\ref{code:helmholtz} after application of padding, data alignment, and \emph{licm}, for an AVX architecture. In this example, sub-expressions invariant to \texttt{j} are identical to those invariant to \texttt{k}, so they can be precomputed once in the $r$ loop.}
\label{code:helmholtz-licm}
\end{algorithm}

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
One notable problem of assembly kernels concerns register allocation and register locality. The critical situation occurs when loop trip counts and the variables accessed are such that the vector-register pressure is high. Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management. Standard optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers is an additional feature of COFFEE. Based on the observation that the evaluation of the element matrix can be reduced to a summation of outer products along the \texttt{j} and \texttt{k} dimensions, a model-driven vector-register tiling strategy can be implemented. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,3
\end{equation}
where $B$ is the set of all basis functions (or temporary variables, e.g. \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Regardless of the specific input problem, by abstracting from the presence of all variables independent of both \texttt{j} and \texttt{k}, the element matrix computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,3$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation. In addition to incurring in larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

\begin{algorithm}[t]
\footnotesize
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// $load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
~~~~~~~~// One $permute\_pd$ intrinsic per \texttt{k}-loop $load$\\
~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsic per \texttt{k}-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz problem after application of \emph{op-vect} on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}. Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length. The original size of the \texttt{j}-\texttt{k} iteration space (i.e. before padding was applied) was 6$\times$6. In this example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\end{algorithm}

% Splitting
\subsection{Expression Splitting}
\label{sec:expr-split}
In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging. If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, poor register reuse is likely to be obtained. This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large. For example, applying loop-invariant code motion to Burgers on a 3D mesh requires 33 temporaries for the \texttt{ijk} loop order. This can make hoisting of the invariant loads out of the \texttt{k} loop inefficient on architectures with a relatively low number of registers. One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}. The transformation can be regarded as a special case of classic loop fission, in which sum's associativity is exploited to distribute the expression across multiple loops. To the best of our knowledge, expression splitting is not supported by available compilers.

@@

\begin{algorithm}[t]
\footnotesize
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~~~// Same code as in Figure~\ref{code:helmholtz-licm} up to line 15 \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz problem in which \emph{split} has been applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
\label{code:helmholtz-split}
\end{algorithm}

@@

\begin{algorithm}[t]
  \textbf{Cost Model}\\
  \KwIn{n\_outer\_arrays, n\_inner\_arrays, n\_consts, n\_regs}
  \KwOut{uaj\_factor, split\_factor}

n\_outer\_regs = n\_regs / 2 \\
split\_factor = 0 \\
// Compute spltting factor \\
\While{n\_outer\_arrays $>$ n\_outer\_regs}{
  n\_outer\_arrays = n\_outer\_arrays / 2 \\
  split\_factor = split\_factor + 1
}

>>>>

\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfig}
\usepackage{tabulary}

\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}

\usepackage{listings}
\lstset{language=C}

\usepackage[cmex10]{amsmath}
\usepackage{url}

@@

@@

\begin{figure}
\begin{alltt}
\footnotesize
\textbf{Input:} element matrix (2D array, initialized to 0), 
       element coordinates (array), 
       coefficient fields (array, e.g. velocity) 
\textbf{Output:} element matrix (2D array)
- Compute Jacobian from coordinates 
- Define basis functions 
- Compute element matrix in an affine loop nest
\end{alltt}
\caption{Structure of a local assembly kernel}
\label{code:general-structure}
\end{figure}

\begin{algorithm}[t]
\SetAlgorithmName{LISTING}{}
\footnotesize
\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$3; k++) \\
~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
~~~~~~~~~~~+((K1*X$\_$D10[i][k]+K3*X$\_$D01[i][k])*(K1*X$\_$D10[i][j]+K3*X$\_$D01[i][j]))+\\
~~~~~~~~~~~+((K0*X$\_$D10[i][k]+K2*X$\_$D01[i][k])*(K0*X$\_$D10[i][j]+K2*X$\_$D01[i][j])))*\\
~~~~~~~~~~~*det*W[i]);\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for a Helmholtz problem on a 2D triangular mesh using Lagrange $p=1$ elements.}
\label{code:helmholtz}
\end{algorithm}

\begin{algorithm}[t]
\SetAlgorithmName{LISTING}{}
\footnotesize
\KwSty{void} burgers(double A[12][12], double **coords, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//11 other basis functions definitions.\\
~~...\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[i][r]);\\
~~~~~~//10 analogous statements (F1, F2, ...)\\
~~~~~~...\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K5*F9)+(K8*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K0*X1$\_$D100[i][k])+(K3*X1$\_$D010[i][k])+(K6*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K2*X2$\_$D100[i][k])+...+(K8*X2$\_$D001[i][k]))*((K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$roughly a hundred sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W[i]);\\
~~$\rbrace$ \\
$\rbrace$
\caption{Local assembly code generated by Firedrake for a Burgers problem on a 3D tetrahedral mesh using Lagrange $p=1$ elements.}
\label{code:burgers}
\end{algorithm}

@@

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions. An example is provided in Listing~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Listing~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse. In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure.

@@

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens of iterations, render the optimization process challenging. It is not always the same set of optimizations that bring performance closest to the machine peak. For example, the Burgers problem in Listing~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Listing~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces. Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.

@@

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From the inspection of the codes in Listings~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions which only depend on two iteration variables. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $intel$ and $gnu$ compilers exhibit. First, they only identify sub-expressions that are invariant with respect to the innermost loop. This is an issue for sub-expressions depending on \texttt{i}-\texttt{k}, which are not automatically lifted in the loop order \texttt{ijk}. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization.

@@

Listing~\ref{code:helmholtz-licm} shows the Helmholtz assembly code after the application of loop-invariant code motion, padding, and data alignment.

\begin{algorithm}[t]
\SetAlgorithmName{LISTING}{}
\footnotesize
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4] ALIGN;\\
~~~~double LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+LI$\_$1[k]*LI$\_$1[j])*det*W[i]);\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code for the Helmholtz problem in Listing~\ref{code:helmholtz} after application of padding, data alignment, and \emph{licm}, for an AVX architecture. In this example, sub-expressions invariant to \texttt{j} are identical to those invariant to \texttt{k}, so they can be precomputed once in the $r$ loop.}
\label{code:helmholtz-licm}
\end{algorithm}

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
One notable problem of assembly kernels concerns register allocation and register locality. The critical situation occurs when loop trip counts and the variables accessed are such that the vector-register pressure is high. Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management. Standard optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers is an additional feature of COFFEE. Based on the observation that the evaluation of the element matrix can be reduced to a summation of outer products along the \texttt{j} and \texttt{k} dimensions, a model-driven vector-register tiling strategy can be implemented. If we consider the code snippet in Listing~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,3
\end{equation}
where $B$ is the set of all basis functions (or temporary variables, e.g. \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Regardless of the specific input problem, by abstracting from the presence of all variables independent of both \texttt{j} and \texttt{k}, the element matrix computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,3$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation. In addition to incurring in larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

\begin{algorithm}[t]
\SetAlgorithmName{LISTING}{}
\footnotesize
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do generalized loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// $load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[0,0], A[1,1], A[2,2], A[3,3] \\
~~~~~~~~// One $permute\_pd$ intrinsic per \texttt{k}-loop $load$\\
~~~~~~~~// Compute A[0,1], A[1,0], A[2,3], A[3,2] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsic per \texttt{k}-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz problem after application of \emph{op-vect} on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}. Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length. The original size of the \texttt{j}-\texttt{k} iteration space (i.e. before padding was applied) was 6$\times$6. In this example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\end{algorithm}

% Splitting
\subsection{Expression Splitting}
\label{sec:expr-split}
In complex kernels, like Burgers in Listing~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging. If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, poor register reuse is likely to be obtained. This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large. For example, applying loop-invariant code motion to Burgers on a 3D mesh requires 33 temporaries for the \texttt{ijk} loop order. This can make hoisting of the invariant loads out of the \texttt{k} loop inefficient on architectures with a relatively low number of registers. One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Listing~\ref{code:helmholtz-split}. The transformation can be regarded as a special case of classic loop fission, in which sum's associativity is exploited to distribute the expression across multiple loops. To the best of our knowledge, expression splitting is not supported by available compilers.

@@

\begin{algorithm}
\SetAlgorithmName{LISTING}{}
\footnotesize
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~~~// Same code as in Listing~\ref{code:helmholtz-licm} up to line 15 \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz problem in which \emph{split} has been applied on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
\label{code:helmholtz-split}
\end{algorithm}

@@
